Forecasting the 2024 US Presidential Election

### 2024-10-31 paper/paper.qmd
Finished model and added final results

---
title: "Forecasting the 2024 US Presidential Election"
subtitle: "My subtitle if needed"
author: 
  - Tina Kim
  - David Flores
  - Kevin Shao
thanks: "Code and data are available at: https://github.com/DavidFJ207/USPresidentialForecast"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(dplyr)
library(knitr)
library(lubridate)
library(ggplot2)
library(ggcorrplot)
library(kableExtra)
library(arrow)
library(zoo)
```

@@

poll_summary <- poll_data %>%
  group_by(state) %>%   
  summarise(
    dem_total_pct = sum(democrat_pct, na.rm = TRUE),
    rep_total_pct = sum(republican_pct, na.rm = TRUE),
    highly_educated_pct = mean(`What is the highest level of education you have attained? College graduate`, na.rm = TRUE),
    lower_educated_pct = mean(`What is the highest level of education you have attained? High school or less`, na.rm = TRUE),
    female_pct = mean(`Can you please tell me your gender? Female`, na.rm = TRUE),
    male_pct = mean(`Can you please tell me your gender? Male`, na.rm = TRUE),
    non_binary_other_pct = mean(`Can you please tell me your gender? Nonbinary or other`, na.rm = TRUE),
    caucasian_pct = mean(`For statistical purposes only, can you please tell me your ethnicity? White or Caucasian`, na.rm = TRUE),
    minority_pct = 100 - caucasian_pct
  ) %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>%  # Replace NA values with 0
  mutate(across(where(is.numeric), ~ round(., 1)))  # Round to nearest tenth

colnames(poll_summary) <- c("State", "Democrat Pct", "Republican Pct", 
                            "High Educated", "Low Educated", "Female", 
                            "Male", "Nonbinary", "Caucasian", 
                            "Minority Ethnicity")

poll_summary_party <- poll_summary %>%
  select(State, `Democrat Pct`, `Republican Pct`)

@@

# Table for party percentages
kable(poll_summary_party) %>%
  kable_styling(font_size = 10, position = "center", full_width = FALSE)

# Second table for demographic variables
kable(poll_summary_demographics) %>%
  kable_styling(font_size = 6, position = "center", full_width = FALSE)

```
We sourced the "Presidential General Election Polls" dataset from FiveThirtyEight [@FiveThirtyEight2024] and performed an in-depth analysis using the statistical programming language R [@citeR].

Our primary objective was to clean, organize, and analyze U.S. presidential election polling data to provide insights into voter preferences by state. This involved selecting data from reputable pollsters, organizing it by state, and addressing any missing or incomplete entries. Additional datasets were merged where necessary, and after thoroughly addressing missing values, we finalized a polished dataset ready for analysis. Below is a table summarizing the polling data by state, showcasing voter preferences by party and demographic breakdowns. The table highlights key trends and differences across states in terms of party lean and demographic distribution. @fig-predictors-pct-summary provides a snapshot of these findings.

## Measurement  
For this analysis, we selected Emerson as our primary pollster. The rationale behind this choice is explained in detail in @sec-appenA. Emerson’s polls provided essential details, including polling date ranges, party affiliations, sample sizes, and the percentage breakdown of support for different political parties.

We focused on analyzing voter preferences based on party affiliation rather than individual candidate support, as this was more relevant to predicting overall trends. By capturing the broader party dynamics, our model aims to forecast how states are likely to swing in future elections, especially with respect to swing states. This focus will be elaborated in our results section, @sec-results.

Additionally, the Emerson dataset included critical demographic data, such as respondents' political preferences, motivations for voting in the 2024 election, opinions on Donald Trump’s legal battles, and perspectives on major political figures and issues. These details, coupled with information on party affiliation, sources of political news, and other personal attributes, allowed us to paint a richer picture of the voting landscape.

To ensure uniformity across all states, we concentrated on questions that were consistently asked across the entire dataset. This enabled us to perform state-level aggregation and analysis without being limited by state-specific polling questions, ensuring a standardized approach across the board.

## Outcome variables
The initial analysis of the dataset revealed some key trends in party preferences across different states. In @fig-party-results, we illustrate which states exhibit the strongest leanings toward either the Democratic or Republican party. This visualization plays a crucial role in identifying the pivotal swing states that lie close to the center of the graph. These swing states, which show near-equal support for both parties, are of particular interest for future election predictions, as they are likely to determine the overall election outcome.

```{r}
#| label: fig-party-results
#| fig-cap: Example Predictor Variable Visualization
#| echo: false

@@

```
```{r}
#| label: fig-party-extremes
#| fig-cap: Demographic Differences Between Top Red and Blue States
#| echo: false
#| warning: false
#| message: false

@@

# Function to identify top N blue and N red states
identify_top_states <- function(data, n) {
  data %>%
    mutate(party_diff = democrat_pct - republican_pct) %>%
    arrange(desc(party_diff)) %>%
    summarise(
      blue_states = list(state[1:n]),  # Use 'n' instead of hardcoding 4
      red_states = list(state[(nrow(data) - n + 1):nrow(data)])
    )
}

# Function to compute mean differences for all numeric variables
compute_mean_differences <- function(data, blue_states, red_states) {
  # Select numeric variables excluding 'state' and 'party_diff'
  numeric_vars <- data %>%
    select(-state, -party_diff) %>%
    select(where(is.numeric)) %>%
    names()

  # Compute means for blue states
  blue_means <- data %>%
    filter(state %in% blue_states) %>%
    summarise(across(all_of(numeric_vars), mean, na.rm = TRUE))

  # Compute means for red states
  red_means <- data %>%
    filter(state %in% red_states) %>%
    summarise(across(all_of(numeric_vars), mean, na.rm = TRUE))

  # Compute absolute differences
  diff_df <- tibble(
    Variable = numeric_vars,
    Blue_Mean = as.numeric(blue_means[1, ]),
    Red_Mean = as.numeric(red_means[1, ]),
    Abs_Difference = abs(as.numeric(blue_means[1, ]) - as.numeric(red_means[1, ]))
  )

  # Remove variables where both means are NA
  diff_df <- diff_df %>%
    filter(!is.na(Blue_Mean) & !is.na(Red_Mean))

@@

# Function to prepare data for plotting
prepare_plot_data <- function(diff_df, top_n = 6) {
  # Get top N variables with the largest differences
  top_diff_vars <- diff_df %>%
    arrange(desc(Abs_Difference)) %>%
    slice(1:top_n)

  # Prepare data for plotting
  plot_data <- top_diff_vars %>%
    select(Variable, Blue_Mean, Red_Mean) %>%
    pivot_longer(
      cols = c("Blue_Mean", "Red_Mean"), 
      names_to = "Group", 
      values_to = "Mean"
    ) %>%
    mutate(Group = recode(Group, "Blue_Mean" = "Blue States", "Red_Mean" = "Red States"))

  plot_data
}

# Function to create the plot with y-axis from 0 to 100 and data labels
create_comparison_plot <- function(plot_data) {
  ggplot(plot_data, aes(
    x = reorder(Variable, -Mean), 
    y = Mean, 
    fill = Group, 
    label = round(Mean, 1)
  )) +
    geom_col(position = position_dodge(width = 0.9)) +
    geom_text(
      position = position_dodge(width = 0.9), 
      vjust = -0.3, 
      size = 1
    ) +
    scale_fill_manual(
      values = c("Blue States" = "blue", "Red States" = "red")
    ) +
    labs(
      title = "Demographic Differences Between Top Red and Blue States",
      x = "Demographic Variable",
      y = "Mean Percentage",
      fill = "State Group"
    ) +
    scale_y_continuous(limits = c(0, 100)) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 4)
    )
}

# Main analysis function
analyze_state_differences <- function(data, top_n_states = 1, top_n_vars = 6) {
  top_states <- identify_top_states(data, n = top_n_states)

@@

```
Beyond identifying swing states, it's also important to understand the differences in demographics between the most strongly Democratic ("blue") and Republican ("red") states. The bar graph in @fig-party-extremes takes a closer look at these differences. It focuses on key factors that show the biggest contrasts between Democratic-leaning and Republican-leaning states, helping to highlight how these factors differ between the most extreme red and blue states.

The @fig-party-extremes visualization provides insight into how people's responses to polls may influence their political preferences. This extra layer of analysis helps clarify the factors shaping the political leanings of different states and adds depth to our understanding of the electorate. In this case, we examine people's opinions on the current president and their voting choices in past elections.

Through these visualizations, we not only identify which states are most likely to swing in future elections but also gain a deeper understanding of the public's opinions that contribute to these partisan divides.

## Predictor variables

Next, we take a deeper dive into the various demographic and social factors that influence voter preferences. The predictors we consider include education levels, gender, ethnicity, and age group. These variables are crucial as they shape political preferences and are likely to influence which way a state leans politically.  
```{r}
#| label: updating-predictors
#| echo: false

@@

In @fig-predictors-party-pct, we explore how these factors correlate with party lean across the U.S. states. For instance, higher levels of education may correlate with a greater likelihood of supporting the Democratic party, while other demographic factors such as ethnicity and age may play different roles in shaping political affiliation. This analysis helps us to identify which factors have the greatest influence in each state, offering a clearer understanding of what drives the electorate.

```{r}
#| label: fig-predictors-party-pct
#| fig-cap: Example Predictor Variable Visualization
#| echo: false

@@

```{r}
#| label: fig-predictors-party-pct-2
#| fig-cap: Example Voting and Approval Predictor Variable Visualization
#| echo: false

# Calculate poll_summary with rounding to the nearest tenth
poll_summary_2 <- poll_data %>%
  group_by(state) %>%
  summarise(
    dem_total_pct = round(sum(.data[[column_refs$democrat]], na.rm = TRUE), 1),
    rep_total_pct = round(sum(.data[[column_refs$republican]], na.rm = TRUE), 1),
    vote_trump_pct = round(mean(.data[[column_refs$vote_trump]], na.rm = TRUE), 1),
    vote_biden_pct = round(mean(.data[[column_refs$vote_biden]], na.rm = TRUE), 1),
    biden_approve_pct = round(mean(.data[[column_refs$biden_approve]], na.rm = TRUE), 1),
    biden_disapprove_pct = round(mean(.data[[column_refs$biden_disapprove]], na.rm = TRUE), 1)
  ) %>%
  mutate(party_diff = round(dem_total_pct - rep_total_pct, 1)) %>%
  drop_na()

@@

```{r}
#| label: fig-predictors-education-race-voting
#| fig-cap: Correlation Matrix of Education, Race, and Past Voting Behavior
#| echo: false

@@

```
The correlation matrix in @fig-predictors-party-pct provides a detailed look at how various predictors are related to party preferences. From this, we can see which predictors—such as education or gender—are more strongly associated with Democratic or Republican leanings. The blue shading indicates stronger correlations with Democratic support, while red represents stronger correlations with Republican support.

```{r}
# Convert `end_date` to Date format and ensure numeric columns
time_data <- time_data %>%
  mutate(end_date = as.Date(end_date, format = "%m/%d/%y"),
         pct = as.numeric(pct), 
         `Do you approve or disapprove of the job Joe Biden is doing as President? Approve` = 
           as.numeric(`Do you approve or disapprove of the job Joe Biden is doing as President? Approve`))

# Average entries for the same date and party, then apply a 14-day rolling average for smoothing
time_data <- time_data %>%
  group_by(end_date, party) %>%
  summarise(
    pct = mean(pct, na.rm = TRUE),
    biden_approval = mean(`Do you approve or disapprove of the job Joe Biden is doing as President? Approve`, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  group_by(party) %>%
  mutate(
    pct_smooth = rollapply(pct, width = 6, FUN = mean, align = "right", fill = NA)
  ) %>%
  ungroup() %>%
  mutate(
    biden_approval_smooth = rollapply(biden_approval, width = 26, FUN = mean, align = "right", fill = NA)
  )

# Separate data for Democrats, Republicans, and Biden Approval
# Filter data for Democrats where state is NA
democrat_data <- time_data %>%
  filter(party == "DEM")

# Filter data for Republicans where state is NA
republican_data <- time_data %>%
  filter(party == "REP")

biden_approval <- time_data %>%
  filter(party == "DEM" | party == "REP") %>%
  select(end_date, biden_approval_smooth) %>%
  distinct()

# Create the time plot with ggplot2
ggplot() +
  # Plot Democrats' smoothed pct over time (blue)
  geom_line(data = democrat_data, aes(x = end_date, y = pct_smooth, color = "Democrats"), size = 1) +

  # Plot Republicans' smoothed pct over time (red)
  geom_line(data = republican_data, aes(x = end_date, y = pct_smooth, color = "Republicans"), size = 1) +

  # Plot Biden's smoothed Approval over time (green)
  geom_line(data = biden_approval, aes(x = end_date, y = biden_approval_smooth, color = "Biden Approval"), size = 1) +

  # Customize labels, theme, colors, and y-axis limits
  labs(title = "Democrat vs. Republican and Biden Approval Over Time (Smoothed)",
       x = "Date",
       y = "Percentage (%)",
       color = "Legend") +
  scale_color_manual(values = c("Democrats" = "blue", "Republicans" = "red", "Biden Approval" = "lightblue")) +
  ylim(25, 55) + # Set y-axis range from 25 to 55
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
This analysis offers valuable insights into how specific demographic factors shape the political landscape in different states, allowing us to better predict electoral outcomes based on shifting demographic trends.

@@

# Model {#sec-model}

## Model Development
The goal of our model is to forecast the popular vote outcome of the 2024 US presidential election.

*Person B:* Choose a linear or generalized linear model. Justify the selection based on your research goal and data structure.

## Model Set-up
Define the model mathematically and contextually (e.g., linear model predicting vote share).

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i x_i \\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

*Person B*: Define all model components and assumptions.

## Model Justification
Our results are summarized in @tbl-modelresults. Discuss the implications of the findings and their predictive accuracy.

>>>>

---
title: "Forecasting the 2024 US Presidential Election"
subtitle: "Analyzing Demographic Patterns and Predicting Swing States"
author: 
  - Tina Kim
  - David Flores
  - Kevin Shao
thanks: "Code and data are available at: https://github.com/DavidFJ207/USPresidentialForecast"
date: today
date-format: long
abstract: Our analysis of swing state polls for the 2024 U.S. Presidential Election shows a close race. Republicans lead in most key states like Michigan, Georgia, Nevada, Wisconsin, and Pennsylvania, while Democrats have a slight edge in Arizona. The national average also leans Republican, suggesting a Republican win in the 2024 Election.
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(dplyr)
library(knitr)
library(lubridate)
library(ggplot2)
library(ggcorrplot)
library(kableExtra)
library(arrow)
library(zoo)
library(brms)
```

@@

# Specify the predictor columns
predictor_columns <- c(
  "What is the highest level of education you have attained? College graduate",
  "What is the highest level of education you have attained? High school or less",
  "Can you please tell me your gender? Female",
  "Can you please tell me your gender? Male",
  "For statistical purposes only, can you please tell me your ethnicity? White or Caucasian",
  "For statistical purposes only, can you please tell me your ethnicity? Hispanic or Latino of any race",
  "Who did you vote for in the 2020 election? Donald Trump",
  "Who did you vote for in the 2020 election? Joe Biden",
  "Do you approve or disapprove of the job Joe Biden is doing as President? Approve",
  "Do you approve or disapprove of the job Joe Biden is doing as President? Disapprove"
)

# Filter poll_data to include only the specified states
poll_summary <- poll_data %>%
  filter(state %in% c("Michigan", "Georgia", "Nevada", "North Carolina", 
                      "New Hampshire", "Wisconsin", "Pennsylvania", "Arizona")) %>%
  group_by(state) %>%   
  summarise(
    dem_total_pct = sum(democrat_pct, na.rm = TRUE),
    rep_total_pct = sum(republican_pct, na.rm = TRUE),
    highly_educated_pct = mean(`What is the highest level of education you have attained? College graduate`, na.rm = TRUE),
    lower_educated_pct = mean(`What is the highest level of education you have attained? High school or less`, na.rm = TRUE),
    female_pct = mean(`Can you please tell me your gender? Female`, na.rm = TRUE),
    male_pct = mean(`Can you please tell me your gender? Male`, na.rm = TRUE),
    non_binary_other_pct = mean(`Can you please tell me your gender? Nonbinary or other`, na.rm = TRUE),
    caucasian_pct = mean(`For statistical purposes only, can you please tell me your ethnicity? White or Caucasian`, na.rm = TRUE),
    minority_pct = 100 - caucasian_pct,
    past_vote_biden = mean(`Who did you vote for in the 2020 election? Joe Biden`, na.rm = TRUE),
    past_vote_trump = mean(`Who did you vote for in the 2020 election? Donald Trump`, na.rm = TRUE)
  ) %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>%  # Replace NA values with 0
  mutate(across(where(is.numeric), ~ round(., 1)))  # Round to the nearest tenth

# Rename columns for clarity
colnames(poll_summary) <- c("State", "Democrat Pct", "Republican Pct", 
                            "High Educated", "Low Educated", "Female", 
                            "Male", "Nonbinary", "Caucasian", 
                            "Minority Ethnicity", "Voted Biden 2020", "Voted Trump 2020")

# Create summary tables
poll_summary_party <- poll_summary %>%
  select(State, `Democrat Pct`, `Republican Pct`, `Voted Biden 2020`, `Voted Trump 2020`)

@@

# Display table for party percentages
kable(poll_summary_party) %>%
  kable_styling(font_size = 10, position = "center", full_width = FALSE)

# Display table for demographic variables
kable(poll_summary_demographics) %>%
  kable_styling(font_size = 6, position = "center", full_width = FALSE)

We sourced the "Presidential General Election Polls" dataset from FiveThirtyEight [@FiveThirtyEight2024] and analyzed it using the statistical programming language R [@citeR].

Our main goal was to clean, organize, and analyze U.S. presidential election polling data to understand voter preferences by state. We selected data from credible pollsters, sorted it by state, and resolved missing or incomplete entries. Where necessary, we merged additional datasets, ensuring that we carefully handled any gaps in the data. The result was a refined dataset ready for analysis. The table below provides a summary of polling data by state, showing voter preferences by party and various demographics. It highlights key trends and differences across states, focusing on party leanings and demographic splits, as illustrated in @fig-predictors-pct-summary.

## Measurement  
We chose Emerson as our primary pollster, and the reasoning for this is detailed in @sec-appenA. Emerson's polls offered crucial information, such as polling dates, party affiliations, sample sizes, and the percentage of support for each political party.

Our analysis centered on voter preferences by party rather than individual candidates. This approach was more relevant for identifying overall trends and helped us forecast which states might swing in future elections, especially swing states. We will discuss this focus further in our results section, @sec-results.

The Emerson dataset also included valuable demographic data, voting history, and approval ratings. To ensure a consistent analysis, we concentrated on questions that were uniformly asked across all states. This approach allowed us to aggregate and analyze data at the state level without being limited by state-specific polling variations.

## Outcome Variables  
Our initial analysis of the dataset revealed important trends in party preferences across states. In @fig-party-results, we show which states have the strongest leanings toward the Democratic or Republican parties. This visualization is crucial for identifying key swing states, which lie near the center of the graph and show nearly equal support for both parties. These states are especially important for predicting the 2024 election outcome.

From this analysis, we identified eight critical swing states: Michigan, Georgia, Nevada, North Carolina, New Hampshire, Wisconsin, Pennsylvania, and Arizona. These states will be central to our election prediction. Although Minnesota emerged as a potential swing state, we excluded it due to the wide range of polling data, as explained in @tbl-final-summary. Additionally, we will consider the national average in our overall prediction.
```{r}
#| label: fig-party-results
#| fig-cap: Example Predictor Variable Visualization
#| echo: false
#| warning: false
#| message: false

@@

```{r}
#| label: tbl-final-summary
#| echo: false
#| warning: false
#| message: false
# Filter time_data for 'DEM' in the 'party' column
dem_data <- time_data %>%
  filter(party == "DEM")

# Calculate 95% confidence interval for each unique state in 'DEM' data
dem_ci_summary <- dem_data %>%
  group_by(state) %>%
  summarise(
    mean_pct = mean(pct, na.rm = TRUE),
    sd_pct = sd(pct, na.rm = TRUE),
    n = sum(!is.na(pct)),
    lower_95 = ifelse(n >= 2, mean_pct - qt(0.975, df = n - 1) * (sd_pct / sqrt(n)), NA),
    upper_95 = ifelse(n >= 2, mean_pct + qt(0.975, df = n - 1) * (sd_pct / sqrt(n)), NA)
  ) %>%
  # Filter out intervals where the bounds go below 40 or above 60, and remove rows where mean_pct exceeds 50
  filter(!(lower_95 < 39 | upper_95 > 60) & mean_pct <= 100)

# Filter time_data for 'REP' in the 'party' column
rep_data <- time_data %>%
  filter(party == "REP")

# Calculate 95% confidence interval for each unique state in 'REP' data
rep_ci_summary <- rep_data %>%
  group_by(state) %>%
  summarise(
    mean_pct = mean(pct, na.rm = TRUE),
    sd_pct = sd(pct, na.rm = TRUE),
    n = sum(!is.na(pct)),
    lower_95 = ifelse(n >= 2, mean_pct - qt(0.975, df = n - 1) * (sd_pct / sqrt(n)), NA),
    upper_95 = ifelse(n >= 2, mean_pct + qt(0.975, df = n - 1) * (sd_pct / sqrt(n)), NA)
  ) %>%
  # Filter out intervals where the bounds go below 40 or above 60, and remove rows where mean_pct exceeds 50
  filter(!(lower_95 < 39 | upper_95 > 60) & mean_pct <= 50)

# Find the intersection of states in both summaries
common_states <- intersect(dem_ci_summary$state, rep_ci_summary$state)

# Filter both summaries to only include the common states
final_dem_ci_summary <- dem_ci_summary %>%
  filter(state %in% common_states)

final_rep_ci_summary <- rep_ci_summary %>%
  filter(state %in% common_states)

# Create the final summary table and round values to 1 decimal point
final_summary_table <- data.frame(
  State = common_states,
  DEM_Mean = round(final_dem_ci_summary$mean_pct, 1),
  DEM_Lower = round(final_dem_ci_summary$lower_95, 1),
  DEM_Upper = round(final_dem_ci_summary$upper_95, 1),
  REP_Mean = round(final_rep_ci_summary$mean_pct, 1),
  REP_Lower = round(final_rep_ci_summary$lower_95, 1),
  REP_Upper = round(final_rep_ci_summary$upper_95, 1)
)

# Print the final summary table using kable
library(knitr)
kable(
  final_summary_table,
  caption = "Final Summary Table of 95% Confidence Intervals for Common States",
  col.names = c("State", "DEM Mean (%)", "DEM Lower 95%", "DEM Upper 95%", "REP Mean (%)", "REP Lower 95%", "REP Upper 95%"),
  format = "markdown"
)

```{r}
#| label: fig-party-extremes
#| fig-cap: Demographic Differences Between Top Red and Blue States
#| echo: false
#| warning: false
#| message: false

@@

# Function to identify top N blue and red states
identify_top_states <- function(data, n) {
  data %>%
    arrange(desc(party_diff)) %>%
    summarise(
      blue_states = list(state[1:n]),  
      red_states = list(state[(nrow(data) - n + 1):nrow(data)])
    )
}

# Function to compute mean differences for all numeric variables
compute_mean_differences <- function(data, blue_states, red_states) {
  numeric_vars <- data %>%
    select(-state, -party_diff) %>%
    select(where(is.numeric)) %>%
    names()

  blue_means <- data %>%
    filter(state %in% blue_states) %>%
    summarise(across(all_of(numeric_vars), mean, na.rm = TRUE))

  red_means <- data %>%
    filter(state %in% red_states) %>%
    summarise(across(all_of(numeric_vars), mean, na.rm = TRUE))

  diff_df <- tibble(
    Variable = numeric_vars,
    Blue_Mean = as.numeric(blue_means[1, ]),
    Red_Mean = as.numeric(red_means[1, ]),
    Abs_Difference = abs(Blue_Mean - Red_Mean)
  ) %>%
  filter(!is.na(Blue_Mean) & !is.na(Red_Mean))

@@

# Function to prepare data for plotting
prepare_plot_data <- function(diff_df, top_n = 6) {
  top_diff_vars <- diff_df %>%
    arrange(desc(Abs_Difference)) %>%
    slice(1:top_n)

  top_diff_vars %>%
    pivot_longer(
      cols = c("Blue_Mean", "Red_Mean"), 
      names_to = "Group", 
      values_to = "Mean"
    ) %>%
    mutate(Group = recode(Group, "Blue_Mean" = "Republican State", "Red_Mean" = "Democrat State"))
}

# Function to create the comparison plot
create_comparison_plot <- function(plot_data) {
  ggplot(plot_data, aes(
    x = reorder(Variable, -Mean), 
    y = Mean, 
    fill = Group, 
    label = round(Mean, 1)
  )) +
    geom_col(position = position_dodge(width = 0.9)) +
    geom_text(
      position = position_dodge(width = 0.9), 
      vjust = -0.3, 
      size = 3
    ) +
    # Update the colors and labels for the groups
    scale_fill_manual(values = c("Republican State" = "red", "Democrat State" = "blue")) +
    labs(
      title = "Demographic Differences Between Top Red and Blue States",
      x = "Demographic Variable",
      y = "Mean Percentage",
      fill = "State Group"
    ) +
    scale_y_continuous(limits = c(0, 100)) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 4)
    )
}

# Main analysis function
analyze_state_differences <- function(data, top_n_states = 1, top_n_vars = 4) {
  top_states <- identify_top_states(data, n = top_n_states)

@@

```
Beyond identifying swing states, it's also important to understand the differences between the strongest Democratic ("blue") and Republican ("red") states. The bar graph in @fig-party-extremes highlights these differences. It focuses on key factors that create the biggest contrasts between Democratic-leaning and Republican-leaning states, helping to show how these factors differ between the most extreme red and blue states.

To start, the graph features two bars, one for Democratic points and one for Republican points, reflecting the responses to questions asked by Emerson Pollster about voting intentions. These responses closely align with the percentages shown for each state, giving us a clear picture of voter preferences.

The @fig-party-extremes visualization gives us valuable insight into how people’s poll responses may influence their political leanings.

Through these visualizations, we not only identify which states are most likely to swing in future elections but also better understand the public opinions driving these partisan divides.

## Predictor Variables

We then take a closer look at the demographic and social factors that shape voter preferences. The predictors we analyze include education levels, gender, and past voting behavior. These variables are important because they significantly impact political leanings and can influence how a state is likely to vote.

```{r}
#| label: updating-predictors
#| echo: false
#| warning: false
#| message: false

@@

In @fig-predictors-party-pct, we examine how these factors correlate with party lean across U.S. states. This analysis helps identify which factors have the greatest influence in each state, providing a clearer understanding of what drives voter behavior.

```{r}
#| label: fig-predictors-party-pct
#| fig-cap: Example Predictor Variable Visualization
#| echo: false
#| warning: false
#| message: false

@@

```{r}
#| label: fig-predictors-party-pct-2
#| fig-cap: Example Voting and Approval Predictor Variable Visualization
#| echo: false
#| warning: false
#| message: false
# Calculate poll_summary with rounding to the nearest tenth
poll_summary_2 <- poll_data %>%
  group_by(state) %>%
  summarise(
    dem_total_pct = round(sum(.data[[column_refs$democrat]], na.rm = TRUE), 1),
    rep_total_pct = round(sum(.data[[column_refs$republican]], na.rm = TRUE), 1),
    vote_trump_pct = round(mean(.data[[column_refs$vote_trump]], na.rm = TRUE), 1),
    vote_biden_pct = round(mean(.data[[column_refs$vote_biden]], na.rm = TRUE), 1),
    biden_approve_pct = round(mean(.data[[column_refs$biden_approve]], na.rm = TRUE), 1),
    biden_disapprove_pct = round(mean(.data[[column_refs$biden_disapprove]], na.rm = TRUE), 1)
  ) %>%
  mutate(party_diff = round(dem_total_pct - rep_total_pct, 1)) %>%
  drop_na()

@@

```{r}
#| label: fig-predictors-education-race-voting
#| fig-cap: Correlation Matrix of Education, Race, and Past Voting Behavior
#| echo: false
#| warning: false
#| message: false

@@

The correlation matrix in @fig-predictors-party-pct shows how different predictors relate to party preferences. We can see which factors, like education or gender, are more strongly tied to Democratic or Republican leanings. Blue shading indicates stronger Democratic support, while red shows stronger Republican support. Key observations include higher education being linked to Democratic support and lower education favoring Trump in 2020. Minority groups generally lean Democratic, while Caucasian/White voters tend to support Trump. Past voter behavior shows Biden supporters remain Democratic, and Trump supporters disapprove of Biden. Additionally, women lean slightly more Democratic compared to men, who often lean Republican.

```{r}
#| label: fig-points-over-time
#| tbl-cap: "Parties Approval"
#| echo: false
#| warning: false
#| message: false
# Convert `end_date` to Date format, ensure numeric columns, and limit to 2024 and onward
time_data_graph <- time_data %>%
  mutate(
    end_date = as.Date(end_date, format = "%Y-%m-%d"),  # Format matches YYYY-MM-DD
    pct = as.numeric(pct),
    biden_approval = as.numeric(`Do you approve or disapprove of the job Joe Biden is doing as President? Approve`)
  ) %>%
  filter(end_date >= as.Date("2024-01-01"))  # Limit data to 2024 and onward

# Separate data for Democrats, Republicans, and Biden Approval
democrat_data <- time_data_graph %>%
  filter(party == "DEM", !is.na(pct)) %>%
  filter(state == "National", !is.na(pct))

republican_data <- time_data_graph %>%
  filter(party == "REP", !is.na(pct)) %>%
  filter(state == "National", !is.na(pct))

biden_approval <- time_data_graph  %>%
  filter(state == "National", !is.na(pct))%>%
  select(end_date, biden_approval) %>%
  filter(!is.na(biden_approval)) %>%
  distinct()

# Create the time plot with ggplot2
ggplot() +
  # Plot Democrats' pct over time (blue)
  geom_line(data = democrat_data, aes(x = end_date, y = pct, color = "Democrats"), size = 1) +

  # Plot Republicans' pct over time (red)
  geom_line(data = republican_data, aes(x = end_date, y = pct, color = "Republicans"), size = 1) +

  # Plot Biden's Approval over time (light blue)
  geom_line(data = biden_approval, aes(x = end_date, y = biden_approval, color = "Biden Approval"), size = 1) +

  # Add a vertical line for a key date
  geom_vline(xintercept = as.Date("2024-07-21"), color = "red", linetype = "dashed") +

  # Customize labels, theme, colors, and y-axis limits
  labs(
    title = "Democrat vs. Republican and Biden Approval Over Time",
    x = "Date",
    y = "Percentage (%)",
    color = "Legend"
  ) +
  scale_color_manual(values = c("Democrats" = "blue", "Republicans" = "red", "Biden Approval" = "lightblue")) +
  ylim(25, 55) + # Set y-axis range from 25 to 55, adjust if needed
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

Lastly, we examine Democratic and Republican support alongside Biden's approval over time, marking July 21st when Joe Biden concluded his campaign, as reported by @globalnews2024. During this period, Democratic support was initially low but increased following this change. By analyzing specific demographic factors and previous voting behavior, we aim to predict the 2024 election outcome.

Key predictor variables:

- **Education**: This measures education levels, with higher education often correlating with more Democratic support.
- **Race/Ethnicity**: This examines how different racial and ethnic groups tend to align politically, with minority groups often leaning Democratic and White/Caucasian voters leaning Republican.
- **Gender**: This captures how gender influences party preference, with women generally leaning more Democratic and men more Republican.
- **Priors (Past Voting Behavior)**: This looks at who people voted for in the past, helping us understand ongoing political support or disapproval.
- **Biden's Approval Rating**: This assesses how approval or disapproval of Joe Biden impacts Democratic support, providing insight into shifting voter sentiments.

@@

# Model {#sec-model}
## Model Development

The goal of our model is to forecast the likely vote outcome in 8 swing states for the 2024 US presidential election. This model will allow us to estimate the expected vote share for each candidate based on state-level characteristics and prior voting patterns, which are especially important in determining close elections.

## Model Set-up

We define the model as a Bayesian linear model where the outcome, \( y_i \), represents the predicted vote share in state \( i \). The predictors in the model capture key demographic and historical voting patterns across swing states, and we aim to model this outcome with an intercept, \( \alpha \), and a state-specific predictor coefficient, \( \beta_i \), that represents the effect of a given variable \( x_i \) on vote share.

### Model Components and Assumptions

- **Outcome Variable:** The predicted vote share (as a percentage) for each candidate in each state.
- **Predictors:** Variables include demographic factors (e.g., education, income, race distribution) and historical voting patterns (e.g., previous election vote shares, approval ratings).
- **Intercept:** Represents the baseline vote share in the absence of additional predictors.
- **Coefficients:** These capture the effect size of each predictor on vote share, allowing for state-specific adjustments.
- **Residual Standard Deviation:** Accounts for unexplained variation in vote share predictions, assumed to follow an exponential distribution.

#### Assumptions
1. **Linearity:** The relationship between each predictor and vote share is assumed to be approximately linear.
2. **Independence:** Vote share predictions for each state are independent, conditional on the predictors.
3. **Normality of Errors:** Errors in vote share predictions are assumed to be normally distributed.
4. **Prior Distributions:** We use weakly informative priors to regularize predictions without imposing strong assumptions.

### Implementation
We run this model in R using the `rstanarm` package, which provides Bayesian estimation with sensible defaults. The default priors in `rstanarm` are chosen to be weakly informative to prevent overfitting while still regularizing our model coefficients.

## Model Justification and Results

By using a Bayesian linear model, we leverage historical and demographic data to provide probabilistic estimates of vote share, giving us a distribution of possible outcomes for each state. This approach allows for uncertainty quantification around predictions, which is particularly important in close elections.

Our results are summarized in Table @tbl-modelresults, which shows the estimated vote shares, posterior means, and credible intervals for each candidate in each swing state. These findings indicate the influence of demographic and historical factors on vote share predictions. Furthermore, the model’s predictive accuracy is evaluated through posterior predictive checks, indicating how well the model captures actual voting patterns. The implications of these findings are significant for forecasting election outcomes, particularly in identifying which states are likely to be more competitive.

```{r}
#| label: fig-results-model
#| tbl-cap: "Forecasting model results based on poll data"
#| echo: false
#| warning: false
#| message: false
# Select relevant columns for the prediction
swing_states <- c("Michigan", "Georgia", "Nevada", "North Carolina", "New Hampshire", "Wisconsin", "Pennsylvania", "Arizona", "National")

# Filter data for the specified swing states and remove any missing data
swing_data <- time_data %>%
  filter(state %in% swing_states, end_date > as.Date("2024-07-21")) %>%
  na.omit()
